The effects of binarity on planet occurrence rates measured by transit surveys

### 2017-11-12 paper/discussion.tex
paper JNW comments round 1

\paragraph{How Has Binarity Been Considered in Occurrence Rate Measurements?}
Many authors have computed planet occurrence rates using transit 
survey data\footnote{
    An online list of occurrence rate papers is maintained at 
    \url{https://exoplanetarchive.ipac.caltech.edu/docs/occurrence_rate_papers.html}
}.
Notable studies using {\it Kepler}\ data include those by Howard et al. 2012, 
Fressin et al., 2013, Foreman-Mackey et al., 2014, Dressing \& Charbonneau 
2015, and Burke et al. 2015.
Mostly if not entirely, these studies have ignored stellar multiplicity.
However, binarity certainly introduces a level of systematic uncertainty to 
pipeline completeness, as well as to star and planet counts.
No one has yet carefully studied the magnitude of this issue for 
{\it Kepler} occurrence rates, and the present work does not resolve this 
problem.
%NOTE: the above might be better in a "discussion" section?

Of course, on a system-by-system level stellar multiplicity affects the 
interpretation of planet candidates. High resolution imaging 
campaigns have consequently been undertaken to measure the multiplicity of 
almost all {\it Kepler}\ Objects of Interest 
(Howell et al. 2011; Adams et al. 2012, 2013; Horch et al. 2012, 2014; 
Lillo-Box et al. 2012, 2014; Dressing et al. 2014; Law et al. 2014; Cartier et 
al. 2015; Everett et al. 2015; Gilliland et al. 2015; Wang et al. 2015a, 
2015b; Baranec et al. 2016).
The results of these programs have been collected by Furlan et al. 
(2017), and they represent an important advance in understanding the KOI 
sample's multiplicity statistics.
In particular, they can be immediately applied to rectify binarity's influence 
on the mass-radius diagram (Furlan \& Howell 2017).

@@

For {\it Kepler}-derived occurrence rates, the high resolution imaging 
campaigns have not yet come full circle to observe a comparison sample of 
non-KOI host stars.
The most recent rate studies have thus used Furlan et al. (2017)'s 
catalog to test the effects of removing KOI hosts with known companions, which 
is an important step towards reducing contamination in the ``numerator'' of 
the occurrence rate (Fulton et al. 2017).
However, without an understanding of the multiplicity statistics of the 
non-KOI hosting stars that are assumed to be searchable, the 
true completeness, the true number of searchable stars, and thus the true 
occurrence rates will remain uncertain.

\paragraph{The Hot Jupiter Rate Discrepancy}
There is at least one context in which measurement of absolute 
occurrence rates may already be showing the signatures of binarity.
Hot Jupiter occurrence rates measured by transit surveys ($\approx 0.5\%$) are 
marginally lower than those found by radial velocity surveys ($\approx 1\%$; 
see Table~\ref{tab:hj_rates}).
Though the discrepancy is of weak statistical significance ($<3\sigma$),
one plausible explanation for the difference is that the populations have 
distinct metallicities.
As originally argued by Gould et al. (2006), the RV sample is biased towards 
metal-rich stars, which have been measured by RV surveys to preferentially 
host more giant planets (Santos et al 2004, Fischer and Valenti 2005).
The {\it Kepler}\ sample specifically has been measured to be more metal poor 
than the local neighborhood, with a mean metallicity of $[{\rm M/H}]_{\rm 
    mean}\approx -0.05$ (Dong et al., 2014; Guo et al., 2017).
Studying the problem in detail, Guo et al. recently argued that the 
metallicity difference could account for a $\approx 0.1\%$ difference in the 
measured rates between the CKS and {\it Kepler}\ samples~--~not a $\approx 
0.5\%$ difference.
Guo et al. concluded that ``other factors, such as binary contamination and 
imperfect stellar properties'' must also be at play.

@@

Ignoring these complications, in this work we focus on whether
binarity might intrinsically bias transit survey occurrence rates, simply 
through its effects on the number of searchable stars and the apparent radii 
of detected planets.

>>>>

\paragraph{How Has Binarity Been Considered in Occurrence Rate Measurements?}
Many authors have computed planet occurrence rates using transit 
survey data\footnote{
    An online list of occurrence rate papers is maintained at 
    \url{https://exoplanetarchive.ipac.caltech.edu/docs/occurrence_rate_papers.html}
}.
Notable studies using {\it Kepler}\ data include those by Howard et al. 2012, 
Fressin et al., 2013, Foreman-Mackey et al., 2014, Dressing \& Charbonneau 
2015, and Burke et al. 2015.
Mostly if not entirely, these studies have ignored stellar multiplicity.
However, binarity clearly introduces a level of systematic uncertainty to 
pipeline completeness, as well as to star and planet counts.
No one has yet carefully studied the magnitude of this issue for 
{\it Kepler} occurrence rates.
While the present work does not resolve this problem, it suggests the 
approximate order of magnitude for the necessary correction factors.
%NOTE: the above might be better in a "discussion" section?

Of course, on a system-by-system level stellar multiplicity affects the 
interpretation of planet candidates. High resolution imaging 
campaigns have been undertaken to measure the multiplicity of 
almost all {\it Kepler}\ Objects of Interest 
(Howell et al. 2011; Adams et al. 2012, 2013; Horch et al. 2012, 2014; 
Lillo-Box et al. 2012, 2014; Dressing et al. 2014; Law et al. 2014; Cartier et 
al. 2015; Everett et al. 2015; Gilliland et al. 2015; Wang et al. 2015a, 
2015b; Baranec et al. 2016).
The results of these programs have been collected by Furlan et al. 
(2017), and they represent an important advance in understanding the KOI 
sample's multiplicity statistics.
In particular, they can be immediately applied to rectify binarity's influence 
on the mass-radius diagram (Furlan \& Howell 2017).

@@

To help clarify {\it Kepler}-derived occurrence rates, the high resolution 
imaging campaigns have not yet come full circle by observing a comparison 
sample of non-KOI host stars.
The most recent rate studies have thus used Furlan et al. (2017)'s 
catalog to test the effects of removing KOI hosts with known companions, which 
is an important step towards reducing contamination in the ``numerator'' of 
the occurrence rate (Fulton et al. 2017).
However, without an understanding of the multiplicity statistics of the 
non-KOI hosting stars that are assumed to be searchable, the 
true completeness, the true number of searchable stars, and thus the true 
occurrence rates will remain uncertain.

\paragraph{The Hot Jupiter Rate Discrepancy}
There is at least one context in which measurement of 
occurrence rates may already be showing the signatures of binarity.
Hot Jupiter occurrence rates measured by transit surveys ($\approx 0.5\%$) are 
marginally lower than those found by radial velocity surveys ($\approx 1\%$; 
see Table~\ref{tab:hj_rates}).
The discrepancy has weak statistical significance ($<3\sigma$).
That said, one reason to expect a difference is that the corresponding stellar 
populations have distinct metallicities.
As argued by Gould et al. (2006), the RV sample is biased towards 
metal-rich stars, which have been measured by RV surveys to preferentially 
host more giant planets (Santos et al 2004, Fischer and Valenti 2005).
The {\it Kepler}\ sample specifically has been measured to be more metal poor 
than the local neighborhood, with a mean metallicity of $[{\rm M/H}]_{\rm 
    mean}\approx -0.05$ (Dong et al., 2014; Guo et al., 2017).
Studying the problem in detail, Guo et al. recently argued that the 
metallicity difference could account for a $\approx 0.1\%$ difference in the 
measured rates between the CKS and {\it Kepler}\ samples~--~not a $\approx 
0.5\%$ difference.
Guo et al. concluded that ``other factors, such as binary contamination and 
imperfect stellar properties'' must also be at play.

@@

Ignoring these complications, in this work we showed that
binarity does bias transit survey occurrence rates, simply through its 
effects on the number of searchable stars and the apparent radii of detected 
planets.
Specifically, our results from Sec.~\ref{sec:model_3} indicate that binarity 
might lead to underestimate HJ rates by a multiplicative factor of $\approx 
1.3$.

We assess the effect this might have towards resolving the hot Jupiter rate 
discrepancy by asking:
what is the probability of Wright et al. (2012)'s result, given a rate drawn 
from the stated bounds of Petigura et al. (in prep)? (See 
Table~\ref{tab:hj_rates}).
In other words, we first take the true HJ rate as $\Lambda_{\rm HJ} = 5.7 \pm 
1.3$, with Gaussian uncertainties, and then drawing from a Poisson 
distribution compute the probability of detecting at least 10 hot Jupiters in 
a sample of 836 stars.
Without accounting for binarity or metallicity, only 4\% of RV surveys would 
detect at least 10 hot Jupiters.
If we multiply $\Lambda_{\rm HJ}$ by $1.2$ to account for Guo et al. 
(2017)'s measured metallicity difference between the {\it Kepler}\ field and 
the local solar neighborhood, 9\% of RV surveys would detect at least 10 hot 
Jupiters.
If we then multiply again by $1.3$ to account for binarity's bias, we find that
23\% of RV surveys would detect at least 10 hot Jupiters, and any discrepancy 
would be quite teneous.
We emphasize that this result is only suggestive~--~a true resolution of the 
rate discrepancy would likely require a detailed understanding of the {\it 
Kepler}\ field's multiplicity statistics.

### 2017-11-12 paper/figures.tex
paper WIP commit
paper JNW comments round 1

\newpage
\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{figures/visualize_volumes.pdf}
    \end{center}
    \caption{
    Cartoon of the different selected and searchable volumes for single 
    stars (left column), and twin binaries (right column).
    This model (\#1) assumes all stars have equal mass and luminosity.
    Since the observer does not recognize the brighter binaries, they select 
    them out to a larger distance, $\sqrt{2}\times$ that of single stars.
    However, dilution causes the twin binaries to only be searchable in one 
    eighth of their selected volume.
    }
    \label{fig:model_1_volumes}
\end{figure}

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{figures/errcases_rate_density_vs_radius_model_1_brokenx.pdf}
    \end{center}
    \caption{
    Inferred planet occurrence rates as a function of planet radius in Model 
    \#1.
    This model has fixed stars, fixed planets, and twin binaries.
    If the true planet radius is $r_p$, all planets 
    detected in binaries will have apparent radii $r_a = r_p/\sqrt{2}$.
    Only 1 in 8 selected binaries is actually searchable (see 
    Sec.~\ref{sec:model_1}).
    To help illustrate the individual effects of errors, we 
    separated them:
    ``only wrong $N_\star$'' means the only error is an incorrectly assumed 
    number of selected stars;
    ``only wrong $Q$'' means the only error is an incorrectly assumed 
    detection efficiency (including both miscalculated transit probabilities 
    and 
    fraction of selected stars that are searchable);
    ``only wrong $r_a$'' means the only error is misinterpretation of 
    planetary radii, due to both transit depth dilution and also wrongly 
    assumed host star radii.
    }
    \label{fig:errcases_model_1}
\end{figure}

\begin{figure}[!tb]
    \centering
    \includegraphics[width=\textwidth]{figures/q_distribn_mag_limited.pdf}
    \caption{
    The distribution of the mass ratio for a magnitude limited sample of 
    binary stars. The underlying mass ratios are drawn from a uniform 
    distribution in a volume-limited sample, quite similar to that of Raghavan 
    et al. (2010)'s Fig. 13.
    The entire bias can be understood analytically (Eq.~\ref{eq:model_2_p_2},
    with $\gamma=0$).
    The numerical comparison uses a realistic empirical mass-luminosity 
    relation rather than $L=M^{3.5}$, found by fitting main sequence dwarf 
    mass luminosity data compiled from Torres et al. [2009] and Benedict et 
    al. [2016].
    }
    \label{fig:q_distribn_mag_limited}
\end{figure}

\begin{figure}[!tb]
    \centering
    \includegraphics[width=.7\textwidth]{figures/errcases_rate_density_vs_radius_model_2.pdf}
    \caption{
    Inferred planet occurrence rates as a function of planet radius in Model 
    \#2.
    This model has fixed planets and primaries, but varying secondaries.
    }
    \label{fig:errcases_model_2_linear}
\end{figure}

\begin{figure}[!tb]
    \centering
    \includegraphics[width=.7\textwidth]{figures/errcases_rate_density_vs_radius_logs_model_2.pdf}
    \caption{
    Same as Fig.~\ref{fig:errcases_model_2_linear}, but with logarithmic 
    $y$-axis, and different $x$ scale.
    }
    \label{fig:errcases_model_2_log}
\end{figure}

\begin{figure}[!tb]
    \centering
    \includegraphics[width=\textwidth]{figures/errcases_rate_density_vs_radius_logs_model_3.pdf}
    \caption{
    Inferred planet occurrence rates as a function of planet radius in Model 
    \#3.
    This model has fixed primaries and single stars, but varying 
    secondaries.
    The true planet radius distribution is a power law with exponent $-2.92$ 
    above $2R_\oplus$, below which it is uniform (e.g., Howard et al., 
    2012).
    }
    \label{fig:errcases_model_3_log}
\end{figure}

\begin{figure}[!tb]
    \centering
    \includegraphics[width=.9\textwidth]{figures/HJ_correction_inputrate_vs_HJratevalues.pdf}
    \caption{
    $\Lambda_i$ is the occurrence rate integrated over all possible phase 
    space for the $i^{\rm th}$ system type. $\Lambda_2/\Lambda_0=1$ 
    corresponds to an equal number of planets per secondary as per single star;
    $\Lambda_2/\Lambda_0=0$ corresponds to secondaries not having any planets.
    In our Model \#3, though the true HJ occurrence rate 
    (Eq.~\ref{eq:occ_rate}) is highly dependent on $\Lambda_2$, 
    the inferred rate hardly depends on whether secondaries have HJs.
    This means that the ``correction factor'' between the inferred rate and 
    the true rate around single stars is underestimated by a multiplicative 
    factor of $\approx1.3$, independent of the HJ rate around secondaries.
    The ``HJ rate'' is the summed rate from     
    Fig.~\ref{fig:errcases_model_3_log} above $8r_\oplus$.
    }
    \label{fig:HJ_correction_inputrate_vs_HJratevalues}
\end{figure}

\begin{figure}[!tb]
    \centering
    \includegraphics[width=.9\textwidth]{figures/earth_inputrate_vs_etaearthratevalues.pdf}
    \caption{
        Same as Fig.~\ref{fig:HJ_correction_inputrate_vs_HJratevalues}, but 
        for Earth-sized planets.
        The absolute values given on the $y$-axis found by summing the rate 
        from Fig.~\ref{fig:errcases_model_3_log} for planetary radii from 
        $0.5$ to $1.5r_\oplus$ (this is a toy model, and they do not reflect 
        an actual determination of $\eta_\oplus$).
        The relative values show that the inferred rate for Earths is roughly 
        independent of the occurrence rate (integrated over all radii) around 
        secondaries.
        However, it is systematically lower than the true rate around single 
        and primary stars, by $\approx 20\%$.
    }
    \label{fig:earth_inputrate_vs_etaearthratevalues}
\end{figure}

>>>>

\newpage

### 2017-11-12 paper/generalities.tex
paper JNW comments round 1

The above definition implicitly marginalizes the rate density over stellar 
multiplicity.
In this study we only consider single and binary star systems.
For a selected population of stars and planets, the rate density is then 
the weighted sum of rate densities for each system type:
\begin{equation}
N_{\rm tot} \Gamma(r) =
N_0 \Gamma_0(r) +
N_1 \Gamma_1(r) +
N_2 \Gamma_2(r),
\label{eq:rate_density_marginalized}
\end{equation}
where $i=0$ corresponds to single stars, $i=1$ to primaries of 
binaries, and $i=2$ to secondaries of binaries.
$N_{\rm tot} = \sum_i N_i$ is the total number of selected stars, and 
$N_0,N_1,N_2$ are the number of selected single stars, primaries, and 
secondaries.
Since each selected binary system contributes both a 
primary and secondary star, $N_1=N_2$.
This redundancy in our notation will later prove its use.

Finally, it is useful to write 
$\Gamma_i(r)$, the rate density for each type of star, as the product of a 
shape function and a constant:
\begin{equation}
\Gamma_i(r) = \frac{{\rm d}\Lambda_i}{{\rm d}r} = Z_i p_i(r),
\quad {\rm for}\ i\in\{0,1,2\}.
\label{eq:rate_density_shape}
\end{equation}
The shape function is normalized to unity.
The $Z_i$'s can be interpreted as each system type's occurrence rate 
$\Lambda_i$, integrated over all planetary radii.
They are equivalent to the number of planets per single star, primary, or 
secondary.

>>>>

The above definition implicitly marginalizes the rate density over stellar 
multiplicity.
In this study we only consider single and binary star systems.
For a selected population of stars and planets, the rate density is then 
a weighted sum of rate densities for each system type:
\begin{equation}
N_{\rm tot} \Gamma(r) =
N_0 \Gamma_0(r) +
N_1 \Gamma_1(r) +
N_2 \Gamma_2(r),
\label{eq:rate_density_marginalized}
\end{equation}
where $i=0$ corresponds to single stars, $i=1$ to primaries of 
binaries, and $i=2$ to secondaries of binaries.
$N_{\rm tot} = \sum_i N_i$ is the total number of selected stars, and 
$N_0,N_1,N_2$ are the number of selected single stars, primaries, and 
secondaries.
Since each selected binary system contributes both a 
primary and secondary star, $N_1=N_2$.
This redundancy in our notation will later prove its use.

Finally, it is helpful to write 
$\Gamma_i(r)$, the rate density for each type of star, as the product of a 
shape function and a constant:
\begin{equation}
\Gamma_i(r) = \frac{{\rm d}\Lambda_i}{{\rm d}r} = Z_i p_i(r),
\quad {\rm for}\ i\in\{0,1,2\}.
\label{eq:rate_density_shape}
\end{equation}
The shape function is normalized to unity.
The $Z_i$'s can be interpreted as each system type's occurrence rate 
$\Lambda_i$, integrated over all planetary radii.
They are equivalent to the number of planets per single star, primary, or 
secondary.

### 2017-11-12 paper/introduction.tex
paper JNW comments round 1

Outside of astrophysical differences, there are observational biases for every 
term in Eq.~\ref{eq:wouldnt_it_be_nice}.
There are errors in $N_{\rm det}$ due to planet radius misclassification.
There are errors in $N_\star$ because a given selected star might in fact be 
two stars.
There are errors in $p_{\rm tra}$ because stars in binaries may have different 
masses and radii than assumed in the single-star case.
Finally, there are errors in $p_{\rm det}$ because the detection efficiencies 
of planets that orbit single stars, primaries, and secondaries are all 
different.

@@

To gain intuition for the many observational biases at play,
we study the following idealized transit surveys:
\begin{itemize}
    \item Model \#1: fixed stars, fixed planets, twin binaries;
    \item Model \#2: fixed planets and primaries, varying secondaries;
    \item Model \#3: fixed primaries, varying planets and secondaries.
\end{itemize}
In Sec.~\ref{sec:numerical_methods}, we introduce our numerical approach 
to the problem, and in Sec.~\ref{sec:analytic_preliminaries} we clarify our 
terminology.
We present the analytic and numerical results in 
Secs.~\ref{sec:model_1}-\ref{sec:model_3}, where each subsection corresponds 
to each model above.
We interpret these calculations throughout, and in 
Sec.~\ref{sec:discussion} discuss their relevance to topical questions in 
the interpretation of transit survey occurrence rates.
In particular, we mention the ``hot Jupiter rate discrepancy'', the relevance 
towards measurements of $\eta_\oplus$, and the significance for the dearth of 
planets recently discovered by Fulton et al. (2017).

>>>>

Outside of astrophysical differences, there are observational biases for every 
term in Eq.~\ref{eq:wouldnt_it_be_nice}.
There are errors in $N_{\rm det}$ due to planet radius misclassification.
There are errors in $N_\star$ because a given selected star might in fact be 
two stars.
There are errors in $p_{\rm tra}$ because stars in binaries may have different 
masses and radii than assumed in the single-star case.
Finally, there are errors in $p_{\rm det}$ because detection efficiencies 
for planets orbiting single stars, primaries, and secondaries are all 
different.

@@

To gain intuition for the many observational biases at play,
we consider the following idealized transit surveys:
\begin{itemize}
    \item Model \#1: fixed stars, fixed planets, twin binaries;
    \item Model \#2: fixed planets and primaries, varying secondaries;
    \item Model \#3: fixed primaries, varying planets and secondaries.
\end{itemize}
We define our terminology in Sec.~\ref{sec:definitions}, and present
our transit survey models in Secs.~\ref{sec:model_1}-\ref{sec:model_3}, where 
subsequent subsections correspond to the models above.
We interpret these calculations throughout, and in 
Sec.~\ref{sec:discussion} discuss their relevance to topical questions in 
the interpretation of transit survey occurrence rates.
In particular, we mention the ``hot Jupiter rate discrepancy'', the relevance 
towards measurements of $\eta_\oplus$, and the significance for the dearth of 
planets recently discovered by Fulton et al. (2017).
We conclude in Sec.~\ref{sec:conclusion}

### 2017-11-12 paper/main.tex
paper JNW comments round 1

This work aims to clarify the biases that stellar binarity introduces
to occurrence rates inferred from transit surveys.
In general, stellar multiplicity leads to diluted planetary 
radii, overestimated detection efficiencies, and an undercounted number of 
selected stars (but possibly an overcounted number of searched stars).
These effects skew occurrence rate measurements in different directions, and 
we develop simple models that allow us to understand the crucial effects.
For models in which all stellar systems are either single or twin binaries, 
and all planets are identical, we find that ignoring binarity leads to a
two-to-threefold underestimate of the occurrence at the true radius.
We believe this represents an upper bound on binarity's effects.
Using more realistic models for the stellar population and planetary 
radii, we find that ignoring binarity leads to a $10$\,-\,$30\%$ 
underestimate of the number of planets per star, depending on the true radius 
distribution.
For the occurrence of Earth-sized planets, in our most realistic model the 
underestimate is by $\approx 15\%$~--~at present far smaller than 
systematic uncertainties on real measurements of this number.
For hot Jupiters, we find that the inferred occurrence rate is 
$\approx 1.3\times$ smaller than the true rate around single stars.
We suggest that this latter effect contributes to the discrepant hot Jupiter 
rates measured by {\it Kepler}\ and the California group's respective surveys.
\end{abstract}

\input{introduction}
\input{methods}
\input{simplest_effects}
\input{results}
\input{discussion}
\input{conclusion}

@@

\input{figures.tex}
\input{hj_rates.tex}

>>>>

This work aims to clarify the biases that stellar binarity introduces
to occurrence rates inferred from transit surveys.
In general, stellar multiplicity leads to diluted planetary 
radii, overestimated detection efficiencies, and an undercounted number of 
selected stars (but possibly an overcounted number of searched stars).
These effects skew occurrence rate measurements in different directions, and 
we develop simple models that allow us to understand the crucial effects.
For a model in which all stellar systems are either single or twin binaries, 
and all planets are identical, we find that ignoring binarity leads to an 
underestimate of the occurrence at the true radius by a multiplicative factor 
of $0.8$ (assuming a 10\% twin binary fraction).
Using more realistic models for the stellar population and planetary 
radii, we find that ignoring binarity leads to a $10$\,-\,$30\%$ 
underestimate of the number of planets per star, depending on the true radius 
distribution.
For the occurrence of Earth-sized planets, in our most realistic model the 
underestimate is by $\approx 15\%$~--~at present far smaller than 
systematic uncertainties on real $\eta_\oplus$ measurements.
For hot Jupiters, we find that the inferred occurrence rate is 
$\approx 1.3\times$ smaller than the true rate around single stars.
We suggest that this latter effect contributes to the discrepant hot Jupiter 
rates measured by {\it Kepler}\ and the California group's respective surveys.
\end{abstract}

\input{introduction}
\input{methods}
\input{results}
\input{discussion}
\input{conclusion}

@@

\input{hj_rates.tex}
